---
title: "COVID-19 in Toronto"
author: "Kevin Li"
date: "3/29/2023"
header-includes:
  - \usepackage{amsmath, amsfonts, amssymb}
  - \usepackage{hyperref}
output: pdf_document
fig_width: 3
fig_height: 2
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
library(tidyverse, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(rattle, warn.conflicts = FALSE)
library(rpart, warn.conflicts = FALSE)
data <- read_csv("COVID19 cases.csv")
```

```{=tex}
\flushleft
\section{Abstract}
```
This report aims to establish a relationship between COVID-19 cases in
Toronto based on data released by the City of Toronto and fatalities,
and will attempt to fit a classifier to determine chance of survival for
a given case. The features that will be analyzed are:

```{=tex}
\begin{itemize}
  \item Age, which is split by decade starting from $\leq 19$ to $\geq 90$
  \item Neighborhood, geographically divided into 140 distinct regions
  \begin{itemize}
    \item Additionally, average income and population by neighbourhood was 
    considered when modeling a relationship - this data 
    was extracted from a different dataset, which can be found here:                            
    \url{https://open.toronto.ca/dataset/neighbourhood-profiles/}
  \end{itemize}
  \item Full Street Address: The first 3 characters of the postal code
  \item Infection source: either travel, household contact close contact, outbreak, 
        or community (none of the former sources)
  \item Classification: either confirmed or probable
  \item Gender
  \item If the patient was ever hospitalized, in ICU, Intubated
  \item Outcome, the feature to be classified - either active, resolved or fatal
\end{itemize}
```
The data was compiled by Toronto Public Health in response to the
COVID-19 pandemic and released under the Open Government License,
available here:
\url{https://open.toronto.ca/dataset/covid-19-cases-in-toronto/}
\newpage
\section{Observations}

Breakdown of cases by year:

```{r echo=FALSE, warning=FALSE}
data <- subset(data, select=-c(`Outbreak Associated`))
rows = nrow(data)
data <- data %>% mutate(`Client Gender` = ifelse(
  `Client Gender` %in% c('NON-BINARY', 'TRANS MAN', 
                         'TRANS WOMAN', 'TRANSGENDER', 'UNKNOWN', 
                         'NOT LISTED, PLEASE SPECIFY'), 'OTHER', `Client Gender`))
data <- data %>% mutate(`Source of Infection` = ifelse(
  `Source of Infection` %in% c('Pending'), 'No Information', `Source of Infection`))
data <- data %>% mutate(`Source of Infection` = ifelse(
  `Source of Infection` %in% c('Outbreaks, Congregate Settings',
                               'Outbreaks, Healthcare Institutions', 
                               'Outbreaks, Other Settings'), 
  'Outbreak', `Source of Infection`))
data <- data %>% mutate(`Ever Hospitalized` = (`Ever Hospitalized`=="Yes")) %>%
  mutate(`Ever in ICU` = (`Ever in ICU`=="Yes")) %>%
  mutate(`Ever Intubated` = (`Ever Intubated`=="Yes"))
data_table <- data %>% group_by(Year=floor_date(`Reported Date`, "year")) %>% 
  mutate(Year = format(Year, "%Y"))
ggplot(data_table, aes(x=Year, fill=`Client Gender`)) + geom_bar(position = "stack") +
  scale_fill_manual(values=c('coral4','steelblue','violet')) + 
  labs(title = "Cases by Gender", y="Total Cases")
ggplot(data_table, aes(x=Year, fill=Outcome)) + geom_bar(position = "fill") +
  scale_fill_manual(values=c('cyan4','rosybrown','greenyellow')) + 
  labs(title = "Cases by Outcome", y="Proportion of Cases")
```

Total fatalities: 4821 (1.22% of cases) \newpage  Rate of patients
admitted to Hospital/ICU/Intubation:

```{r echo=FALSE, warning=FALSE, message=FALSE}
patients_hosp <- data %>% group_by(`Ever Hospitalized`, 
                                   `Ever in ICU`, `Ever Intubated`) %>%
  summarize(N = n(), Proportion=n()/rows)
patients_hosp
```

Patients by source of infection:

```{r echo=FALSE, warning=FALSE, message=FALSE}
patients_source <- data %>% group_by(`Source of Infection`) %>%
  summarize(N = n(), Proportion=n()/rows)
patients_source
```

Average age of infection: since the ages of patients are released by age
range, calculating the average age of infection can be done using a
confidence interval. A random sample of size 1000 will be taken from the
dataset to reduce the running time of the bootstrap function. Assuming
that patients within a certain age range (ex. 50-59 years) are
distributed uniformly, i.e. $X_{a-b} \sim Uniform(a, b)$, a random age
will be assigned to each patient accordingly using \verb+runif+. For the
age range of 90+ years, an exponential distribution will be used
instead($X_{90+} \sim exp(1) + 90$). A bootstrap confidence interval can
then be established using the random ages, which will accurately
represent the true random age.\
Mean age of patient:

```{r echo=FALSE, warning=FALSE, message=FALSE}
age <- sample_n(subset(data, select=c(`Age Group`)), 1000)
age <- na.omit(age)
rand_age = function(name){
  if(str_equal(name[1], '19 and younger')){
    return(runif(1, min=0, max=19)[1])
  }
  if(str_equal(name[1], '90 and older')){
    return(rexp(1)[1]+90.0)
  }
  a = as.numeric(substr(name[1], 1, 2))[1]
  b = as.numeric(substr(name[1], 7, 8))[1]
  return(runif(1, min=a, max=b))
}
random_age <- apply(age, 1, rand_age)
boot = function(){
  boot_sam = sample(random_age, size=1000, replace=TRUE)
  return(mean(boot_sam))
}
boot_mean <- replicate(1000, boot())
quantile(boot_mean, c(0.025, 0.975))
```

\newpage

The relationship between income and number of cases can be determined by
fitting a linear model between the average income by neighbourhood and
the proportion of cases to total population.\
The test hypothesis of the linear model is:
$H_0: Income \perp \frac{cases}{population}$\
while the alternative hypothesis is:
$H_a: Income \not\perp \frac{cases}{population}$

```{r echo=FALSE, warning=FALSE, message=FALSE}
data_neighbour <- data %>% group_by(`Neighbourhood Name`) %>% 
  summarize(N = n()) %>% head(n=140)

neighbour_prof <- read_csv("neighbourhood-profiles-2016-140-model.csv")
average_income <- neighbour_prof[945,] %>% t() %>% 
  tail(n=140)
average_income_thousands <- as.numeric(gsub(",", "", average_income))/1000
total_pop <- neighbour_prof[3,] %>% t() %>% tail(n=140)
total_pop <- as.numeric(gsub(",", "", total_pop))
data_neighbour <- data_neighbour %>% mutate(Prop_cases_pop = N/total_pop*100)
lin_model <- lm(data_neighbour$Prop_cases_pop~average_income_thousands)
summary(lin_model)
plot(average_income_thousands, data_neighbour$Prop_cases_pop, col="blue", pch=1, cex=0.5, 
     main="Wealth Inequality and COVID-19 in Toronto", cex.main=0.8, 
     xlab="Average income per Thousand", 
     ylab="% of Cases Relative to Total Population", cex.lab=0.8)
abline(lin_model)
```

The intercept coefficient of 15 implies that when average income is 0,
the proportion of cases to neighbourhood population is 15%. The
average_income coefficient of -0.07 reveals that the rate of cases drops
by 7% per \$1000 in additional average income.\
A p-value of 0.159 and R-squared value of 0.014 imply that the
correlation between average income and rate of cases by neighbourhood is
extremely weak. There is insufficient evidence to reject the null
hypothesis of income being unrelated to the probability of contracting
COVID-19. \newpage Similarly, the relationship between average income
and chance of fatality per case can be fitted with a linear model using
the following hypotheses:
$H_0: Income \perp \frac{Fatalities}{Total Cases}$\
$H_a: Income \not\perp \frac{Fatalities}{Total Cases}$

```{r echo=FALSE, warning=FALSE, message=FALSE}
data_neighbour_fatal <- data  %>% filter(str_equal(`Outcome`, "FATAL")) %>% 
  group_by(`Neighbourhood Name`) %>% summarize(N = n()) %>% head(n=140)
data_neighbour_fatal <- data_neighbour_fatal$N/data_neighbour$N * 100
lin_model_fatal <- lm(data_neighbour_fatal~average_income_thousands)
summary(lin_model_fatal)
plot(average_income_thousands, data_neighbour_fatal, col="red", pch=1, cex=0.5, 
     main="Wealth Inequality and COVID-19 Fatalities in Toronto", cex.main=0.8, 
     xlab="Average income per Thousand", 
     ylab="% of Fatal Cases Relative to Total Cases", cex.lab=0.8)
abline(lin_model_fatal)
```

An intercept coefficient of 1.32 implies that when the average income is
\$0, the fatality of cases is approximately 1.32%. The
average_income_thousands coefficient of -0.006 implies that the
lethality of COVID-19 drops by around 0.6% per additional thousand
dollars of average income.\
The correlation is even lower than before, signalling that there is
almost no relationship between fatality rate and average income of the
patient. Likewise, a p-value of almost 0.5 does not provide sufficient
evidence to reject the null hypothesis of fatality rate being unrelated
to income. 
\newpage
Finally, a 4-fold cross validation technique will be 
implemented to develop a classifier. The features used for the classifier will
be:
\begin{itemize}
  \item Age Group
  \item Source of Infection
  \item Client Gender
  \item Ever Hospitalized/in ICU/Intubated
\end{itemize}
It is likely that source of infection and gender may become statistically irrelevant
as the information gained from these features is very small. 
```{r echo=FALSE, warning=FALSE, message=FALSE}
data <- data %>% mutate(`Outcome` = ifelse(str_equal(`Outcome`, "FATAL"), "Fatal", "Not fatal")) %>%
  mutate(group_ind = sample(c(1,2,3,4), size=nrow(data), replace=T)) %>% na.omit()
names(data) <- make.names(names(data))
accuracy_vec <- vector()
for(i in 1:4){
  data_train <- data %>% filter(group_ind != i)
  Dec_tree <- rpart(`Outcome` ~ `Age.Group` + 
                    `Source.of.Infection` + `Client.Gender` + 
                    `Ever.Hospitalized` + `Ever.in.ICU` + `Ever.Intubated`,  
                    data=data_train, method = "class")
  data_test <- data %>% filter(group_ind == i)
  result <- predict(Dec_tree, data_test, type="class")
  conf_mat <- table(result, data_test$`Outcome`)
  accuracy_vec[i] <- sum(diag(conf_mat))/sum(conf_mat)
  if(i == 1){
    fancyRpartPlot(Dec_tree)
  }
}
print("4-fold cross-validation results: ")
print(accuracy_vec)
print("Accuracy of decision tree classifier: ")
print(mean(accuracy_vec))
```

\newpage
\section{Conclusion}  
COVID-19 has had detrimental effects in the city of Toronto. While the overall
fatality rate of the disease is lower than its predecessor SARS, its overwhelming
prevalence has resulted in many more deaths. COVID is mostly benign, since
under 5% of confirmed or suspected infections require hospitalization. The number
of infections between women and men was approximately equal in 2020 and 2021, but more
women were infected than men in 2022. The mean age of infection in Toronto is likely 
somewhere between 40 and 43, and a weak negative
correlation between average income and total cases could be found - but there was
no evidence to suggest fatalities were linked to income. Finally, decision tree
analysis reveals that intubated patients over the age of 60 are more likely
to die, while intubated patients under the age of 60 were more likely to survive. 

\newpage

```{r eval=FALSE, warning=FALSE, message=FALSE}
# Initial setup
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
library(tidyverse, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(rattle, warn.conflicts = FALSE)
library(rpart, warn.conflicts = FALSE)
data <- read_csv("COVID19 cases.csv")

# Basic information (Client cases by Gender/Outcome)
data <- subset(data, select=-c(`Outbreak Associated`))
rows = nrow(data)
data <- data %>% mutate(`Client Gender` = ifelse(
  `Client Gender` %in% c('NON-BINARY', 'TRANS MAN', 
                         'TRANS WOMAN', 'TRANSGENDER', 'UNKNOWN', 
                         'NOT LISTED, PLEASE SPECIFY'), 'OTHER', `Client Gender`))
data <- data %>% mutate(`Source of Infection` = ifelse(
  `Source of Infection` %in% c('Pending'), 'No Information', `Source of Infection`))
data <- data %>% mutate(`Source of Infection` = ifelse(
  `Source of Infection` %in% c('Outbreaks, Congregate Settings',
                               'Outbreaks, Healthcare Institutions', 
                               'Outbreaks, Other Settings'), 
  'Outbreak', `Source of Infection`))
data <- data %>% mutate(`Ever Hospitalized` = (`Ever Hospitalized`=="Yes")) %>%
  mutate(`Ever in ICU` = (`Ever in ICU`=="Yes")) %>%
  mutate(`Ever Intubated` = (`Ever Intubated`=="Yes"))
data_table <- data %>% group_by(Year=floor_date(`Reported Date`, "year")) %>% 
  mutate(Year = format(Year, "%Y"))
ggplot(data_table, aes(x=Year, fill=`Client Gender`)) + geom_bar(position = "stack") +
  scale_fill_manual(values=c('coral4','steelblue','violet')) + 
  labs(title = "Cases by Gender", y="Total Cases")
ggplot(data_table, aes(x=Year, fill=Outcome)) + geom_bar(position = "fill") +
  scale_fill_manual(values=c('cyan4','rosybrown','greenyellow')) + 
  labs(title = "Cases by Outcome", y="Proportion of Cases")

# Rate of Hospitalization
patients_hosp <- data %>% group_by(`Ever Hospitalized`, 
                                   `Ever in ICU`, `Ever Intubated`) %>%
  summarize(N = n(), Proportion=n()/rows)
patients_hosp

# Source of transmission
patients_source <- data %>% group_by(`Source of Infection`) %>%
  summarize(N = n(), Proportion=n()/rows)
patients_source

# Mean age
age <- sample_n(subset(data, select=c(`Age Group`)), 1000)
age <- na.omit(age)
rand_age = function(name){
  if(str_equal(name[1], '19 and younger')){
    return(runif(1, min=0, max=19)[1])
  }
  if(str_equal(name[1], '90 and older')){
    return(rexp(1)[1]+90.0)
  }
  a = as.numeric(substr(name[1], 1, 2))[1]
  b = as.numeric(substr(name[1], 7, 8))[1]
  return(runif(1, min=a, max=b))
}
random_age <- apply(age, 1, rand_age)
boot = function(){
  boot_sam = sample(random_age, size=1000, replace=TRUE)
  return(mean(boot_sam))
}
boot_mean <- replicate(1000, boot())
quantile(boot_mean, c(0.025, 0.975))

# Regression: average income and COVID-19 cases 
data_neighbour <- data %>% group_by(`Neighbourhood Name`) %>% 
  summarize(N = n()) %>% head(n=140)

neighbour_prof <- read_csv("neighbourhood-profiles-2016-140-model.csv")
average_income <- neighbour_prof[945,] %>% t() %>% 
  tail(n=140)
average_income_thousands <- as.numeric(gsub(",", "", average_income))/1000
total_pop <- neighbour_prof[3,] %>% t() %>% tail(n=140)
total_pop <- as.numeric(gsub(",", "", total_pop))
data_neighbour <- data_neighbour %>% mutate(Prop_cases_pop = N/total_pop*100)
lin_model <- lm(data_neighbour$Prop_cases_pop~average_income_thousands)
summary(lin_model)
plot(average_income_thousands, data_neighbour$Prop_cases_pop, col="blue", pch=1, cex=0.5, 
     main="Wealth Inequality and COVID-19 in Toronto", cex.main=0.8, 
     xlab="Average income per Thousand", 
     ylab="% of Cases Relative to Total Population", cex.lab=0.8)
abline(lin_model)

# Regression: average income and fatal cases
data_neighbour_fatal <- data  %>% filter(str_equal(`Outcome`, "FATAL")) %>% 
  group_by(`Neighbourhood Name`) %>% summarize(N = n()) %>% head(n=140)
data_neighbour_fatal <- data_neighbour_fatal$N/data_neighbour$N * 100
lin_model_fatal <- lm(data_neighbour_fatal~average_income_thousands)
summary(lin_model_fatal)
plot(average_income_thousands, data_neighbour_fatal, col="red", pch=1, cex=0.5, 
     main="Wealth Inequality and COVID-19 Fatalities in Toronto", cex.main=0.8, 
     xlab="Average income per Thousand", 
     ylab="% of Fatal Cases Relative to Total Cases", cex.lab=0.8)
abline(lin_model_fatal)

# Classifier: Fatal cases
data <- data %>% mutate(`Outcome` = ifelse(str_equal(`Outcome`, "FATAL"), "Fatal", "Not fatal")) %>%
  mutate(group_ind = sample(c(1,2,3,4), size=nrow(data), replace=T)) %>% na.omit()
names(data) <- make.names(names(data))
accuracy_vec <- vector()
for(i in 1:4){
  data_train <- data %>% filter(group_ind != i)
  Dec_tree <- rpart(`Outcome` ~ `Age.Group` + 
                    `Source.of.Infection` + `Client.Gender` + 
                    `Ever.Hospitalized` + `Ever.in.ICU` + `Ever.Intubated`,  
                    data=data_train, method = "class")
  data_test <- data %>% filter(group_ind == i)
  result <- predict(Dec_tree, data_test, type="class")
  conf_mat <- table(result, data_test$`Outcome`)
  accuracy_vec[i] <- sum(diag(conf_mat))/sum(conf_mat)
  if(i == 1){
    fancyRpartPlot(Dec_tree)
  }
}
print("4-fold cross-validation results: ")
print(accuracy_vec)
print("Accuracy of decision tree classifier: ")
print(mean(accuracy_vec))
```
